### 2020-12-17 paper/sex_selection_spacing-ver5.tex
Specify birth interval length when needed

I use the model to predict birth interval measures, parity progression probabilities, 
and the sex ratio rather than present coefficients because the interpretation 
of competing risk model coefficients is challenging \citep{thomas96}.
The predicted parity progression probability is the likelihood of giving birth by the
imposed censoring based on standard survival curve calculations averaged across all
women in a given sample.

For birth intervals, I estimate a set of percentile birth intervals. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% mark on her survival 
curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

In addition to a large number of women surveyed and the long period covered, a significant 
benefit of the NFHS over other surveys is that enumerators pay careful attention to the 
spacing between births and probe for ``missed'' births.
For India, the main concern is underreporting of deceased children, especially a systematic 
recall error where respondents' likelihood of reporting the birth of a deceased child depends 
on the sex of that child.
Unreported deceased children inflate the birth intervals and, with declining mortality, 
make changes over time appear too small.
In the online appendix, I provide a detailed analysis of systematic recall error, which 
shows that recall error depends heavily on how long ago a woman was married. 
I, consequently, drop women married 22 years or more.%
\footnote{
Recall error is likely behind the designation of the first two rounds of NFHS as ``moderate 
quality'' in an analysis of the quality of birth histories in DHS surveys and its impact 
on fertility estimates \citep{Schoumaker2014}.
}

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentile birth intervals in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with 12 or more years of 
education, and the fourth spell for women with 12 or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2\% boys
\citep{ben-porath76b,jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors are available in the online 
appendix.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with no education by spell, sex composition, and period}
\label{fig:spacing_low_rural}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with 1--7 years of education by spell, sex composition, and period}
\label{fig:spacing_med_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with 1--7 years of education by spell, sex composition, and period}
\label{fig:spacing_med_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with 8--11 years of education by spell, sex composition, and period}
\label{fig:spacing_high_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with 8--11 years of education by spell, sex composition, and period}
\label{fig:spacing_high_urban}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with 12 or more years of education by spell, sex composition, and period}
\label{fig:spacing_highest_urban}
\end{figure}

@@

% [Birth intervals w/o sex selection]
\subsection{When Sex Selection Is Less Used}

To separate the effects of the introduction of sex selection and the other changes in
India, I first discuss how birth intervals have changed in situations where sex selection 
is less used.
The group broadly covers women with no education, regardless of the sex composition of their 
children, and women with any education who have one or two sons already. 
Despite the lower level of sex selection, son preference is still evident with the 
shortest spacing when they have only girls. 
Notably, for those least likely to use sex selection---rural women with no education---the 
difference in birth intervals across sex compositions has grown over time as spacing when 
sons are present has increased.

% [25th percentile birth intervals]
A remarkably high proportion of birth intervals are still very short.
For all but the most educated, 25\% or more have their second and third child 
within 24 months of the previous birth.
These intervals are substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Furthermore, despite higher parities' more substantial increases in birth intervals, even 
the 25th percentile birth intervals for the fourth spell are around 24 months for women 
with less than eight years of education.

% [median birth intervals]
Median birth intervals have also increased relatively little---only three to six months 
over the four decades---compared to around 3.5 months \emph{per decade} in other countries with 
declining fertility \citep{Rutstein2011,Casterline2016}.%
\footnote{
The NFHS reports show median closed birth intervals of approximately 31 months, which have 
barely moved over time, underscoring the importance of accounting for censoring when examining 
birth spacing.
}
The result is that most of the median birth intervals are still at 36 months or below, 
with the shortest only 29 months.

Birth intervals appeared to lengthen the most for women the least likely to work.
For example, from lowest to highest education, the average third-spell birth intervals for 
urban women with one boy and one girl increased by 2.7, 3.4, 5.8, and 1.8 months over the 
four decades.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0.
See the online appendix tables for the average birth intervals.
}
Hence, women with the lowest labor force participation---those with 8--11 years 
of education---also saw the largest increases in average spacing, possibly driven by the 
substantial improvement in household income for this group from economic growth.

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth intervals, where the 
more the parity progression probabilities declined, the more the birth interval lengthened. 
For example, the probability of a fourth birth for urban women with 8--11 years 
of education and two sons and a girl has declined by almost 40 percentage points as the 
75th percentile birth interval increased by 22 months.
Compare this with rural women of no education with a boy as their first child, for whom the 
probability of a third birth declined by fewer than six percentage points while the birth 
interval increased only slightly over two months.

@@

These results are in line with prior research showing that falling fertility is associated 
with increases in longer spacing, although why is still an unresolved question 
\citep{Casterline2016}.
The exception to this trend suggests one possible answer.
For the most educated women who already have a son, the probability of a third birth 
declined rapidly, but the birth intervals changed little.
These women both have better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}.

% [Interpretation]
% How do these changes in birth intervals tie in with the changes in India over the four
% decades?
% [TK economic growth? -> higher return to education for males -> lower fertility and
% higher investment in boys' education -- has anybody looked at this?]
% First, the parity progression probabilities have been falling over time across all 
% education levels, especially for women with at least one son.

@@

A clear illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
With two girls, almost 80\% of the third births are boys, and the 75th percentile 
birth interval is close to 70 months. 
This interval is about 13 months longer than if they had at least one son and represents 
an increase of almost 21 months over the four decades.

Even more striking is that most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These changes in birth spacing may even be an underestimate because this particular group 
appears to have had access to sex selection even before it became widespread, as shown by 
the unequal sex ratio for the 1972--1984 period for women with two girls.

The 75th percentile changes are the most dramatic, but sex selection also affects the 
25th and median birth intervals. 
For the best-educated urban women with two girls, the 25th percentile birth interval 
increased by six months, or 23\%, while the median percentile birth interval 
increased by 15 months (43\%).

@@

% An important exception to the relatively short median intervals come from the better 
% educated women, especially in the absence of sons.
% The longest median interval was 48 months for the third spell for urban women with twelve 
% or more years of education who had no boys, an increase of almost 15 months over
% four decades.
% Similary, urban women with eight to eleven years of education and no sons had median
% birth intervals of around 40 months, an increase of about 10 months.
% Finally, the best educated women now have a median second birth interval of around 42
% months whether their first child was a boy or a girl.

% Despite the strong son preference in spacing and sex ratios, there is also some indication 
% that parents prefer a mix of boys and girls;
% women with two boys \emph{and} a girl consistently have the lowest parity progression 
% probability in the fourth spell across education levels.

@@

Sex selection, however, is behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth intervals. 
Over the four decades, the median birth interval for this group increased by almost 15 
months, 
and the 75th percentile birth interval increased by a staggering 21 months, most of that 
within a decade.

@@

Consistent with a more substantial bias in the fertility rate when the age of marriage 
and the length of birth intervals increase, the absolute bias is least in the first 
and the last period and highest in the middle two periods.
Hence, the fertility rate declined too fast from the mid-1980s to the century's end.
Only recently, as the rate of increase for the birth intervals has slowed, have the two 
fertility measures begun to converge again.
Even with the convergence, the predicted 2005--2016 cohort fertility is still above the 
1992--1993 fertility rate for every group, except urban women with no education. 
Furthermore, for the last period, the predicted cohort fertility remains at least 
10\%--20\% higher than the fertility rate.

@@

Finally, even with the declines in the predicted cohort fertility, it is still mostly 
above replacement.
Only for urban women with 12 or more years of education is the predicted cohort
fertility clearly below 2.1 children.
Even then, cohort fertility is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Furthermore, the predicted cohort fertility numbers are likely too low because I use only 
the first four births and births before the imposed 105-month birth interval censoring.

@@

The dependent variable is whether the child died within the first 12 months of life.
The main set of explanatory variables consists of dummies for the spacing from
the prior birth.
The birth interval dummies cover 12-month periods, starting nine months after the prior 
birth, until the 57-month dummy, which covers until 105 months after the prior birth.
I use dummies for sex of the index child and the sex composition 
of the prior children.
The birth spacing dummies, the sex of index child, and the sex composition dummies
are all interacted.
Because the actual number of abortions is unobserved, the interactions between
the sex composition of prior children and the sex of the index child serve
as proxies for the use of sex selection.
The other explanatory variables are the same as above, and estimations
are done separately by education level and parity.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_2_low_med}
\caption{Infant mortality by preceding birth interval across periods for second child of women with 
no education and women with 1--7 years of education}
\label{fig:mortality_low_med}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_2_high_highest}
\caption{Infant mortality by preceding birth interval across periods for second child of women with 
8--11 and 12 and above years of education}
\label{fig:mortality_high_highest}
\end{figure}

@@

% What has happened to spacing and why
The most substantial lengthening of birth intervals came from the best-educated women
because of their substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio increased to close to 80\% boys, the expected median birth interval 
increased by almost 15 months, and the 75th percentile interval increased by 21 months.
Most of the increase in the long intervals came immediately after the introduction of sex 
selection in India.

@@

Tempo effects are studied extensively in the literature \citep[see, for example, ][]{Bongaarts1999}.
Still, there are, to my knowledge, no other cases where there has been as 
substantial an increase in birth intervals and associated bias in fertility rates 
as for India.
It is conceivable that we might see increases in the total fertility rate as birth spacing 
stabilizes or even shortens again if interventions against sex selection are successful.

@@

NFHS enumerators probe for any missed births, although the method depends on the survey.
NFHS-1 probe for each calendar birth interval that is four or more years.
NFHS-2 asked for stillbirths, spontaneous and induced abortions and also probed 
for each calendar birth interval four or more years.
NFHS-3 and NFHS-4 did not directly use birth intervals, but asked whether there were any 
other live births between (name of previous birth) and (name), including any children who 
died after birth, and asked for births before the birth listed as first birth and
after the last birth listed as the last birth.

@@

Figures \ref{fig:spacing_low_urban} and \ref{fig:spacing_highest_rural} show 25th, 50th,
and 75th percentile birth intervals, the sex ratio, and the probability of parity
progression by spell for urban women with no education and rural women with 12 or
more years of education, respectively.

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_urban}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for urban women with no education by spell, sex composition, and period}
\label{fig:spacing_low_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_rural}
\caption{Percentile birth intervals, sex ratios, and parity progression  
for rural women with 12 or more years of education by spell, sex composition, and period}
\label{fig:spacing_highest_rural}
\end{figure}

The first set of tables, Tables \ref{tab:p25_p50_p75_low}, \ref{tab:p25_p50_p75_med},
\ref{tab:p25_p50_p75_high}, and \ref{tab:p25_p50_p75_highest} show 25th, 50th, and
75th percentile birth intervals together with their standard errors.
The standard errors for all measures are based on bootstrapping,
where the model is repeatedly estimated using resampling with replacement.

The second set of tables, 
Tables \ref{tab:avg_sex_ratio_low}, \ref{tab:avg_sex_ratio_med}, 
\ref{tab:avg_sex_ratio_high}, and \ref{tab:avg_sex_ratio_highest}, show predicted average 
birth intervals, sex ratios, and probabilities of having a birth by decade, spell, and sex 
composition for the four education levels separated by the area of residence, together 
with bootstrapped standard errors for all three outcomes.
To find the average birth interval, I calculate, for each woman, the probability of 
giving birth in each $t$, and her expected spell length from these probabilities. 
I then average the individual expected spell lengths across women using their parity 
progression probabilities as weights. Finally, I add nine months because spells begin 
nine months after the previous birth.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_3_low_med}
\caption{Infant mortality by preceding birth interval across periods for third child of women with 
no education and women with 1--7 years of education}
\label{fig:mortality_low_med_spell_3}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_3_high_highest}
\caption{Infant mortality by preceding birth interval across periods for third child of women with 
8--11 and 12 and above years of education}
\label{fig:mortality_high_highest_spell_3}
\end{figure}

>>>>

I use the model to predict birth interval lengths, parity progression probabilities, 
and the sex ratio rather than present coefficients because the interpretation 
of competing risk model coefficients is challenging \citep{thomas96}.
The predicted parity progression probability is the likelihood of giving birth by the
imposed censoring based on standard survival curve calculations averaged across all
women in a given sample.

For birth interval lengths, I estimate a set of percentile durations. 
I first calculate for each woman when there is a given percentage chance that she will have 
given birth, conditional on the probability of giving birth in the spell. 
For example, with an 80\% parity progression probability, the median birth interval 
is the predicted number of months before a woman passes the 60\% mark on her survival 
curve. 
I then add nine months to account for the start of the spell. 
The reported statistic is the average of a given percentile interval across all women in 
a given sample using the individual progression probabilities as weights.

@@

In addition to a large number of women surveyed and the long period covered, a significant 
benefit of the NFHS over other surveys is that enumerators pay careful attention to the 
spacing between births and probe for ``missed'' births.
For India, the main concern is the underreporting of deceased children, especially a systematic 
recall error where respondents' likelihood of reporting the birth of a deceased child depends 
on the sex of that child.
Unreported deceased children inflate the length of birth intervals and, with declining mortality, 
make changes over time appear too small.
In the online appendix, I provide a detailed analysis of systematic recall error, which 
shows that recall error depends heavily on how long ago a woman was married. 
I, consequently, drop women married 22 years or more.%
\footnote{
Recall error is likely behind the designation of the first two rounds of NFHS as ``moderate 
quality'' in an analysis of the quality of birth histories in DHS surveys and its impact 
on fertility estimates \citep{Schoumaker2014}.
}

@@

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentile birth interval durations in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with 12 or more years of 
education, and the fourth spell for women with 12 or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2\% boys
\citep{ben-porath76b,jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors are available in the online 
appendix.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_rural}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for rural women with no education by spell, sex composition, and period}
\label{fig:spacing_low_rural}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_rural}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for rural women with 1--7 years of education by spell, sex composition, and period}
\label{fig:spacing_med_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_med_urban}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for urban women with 1--7 years of education by spell, sex composition, and period}
\label{fig:spacing_med_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_rural}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for rural women with 8--11 years of education by spell, sex composition, and period}
\label{fig:spacing_high_rural}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_high_urban}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for urban women with 8--11 years of education by spell, sex composition, and period}
\label{fig:spacing_high_urban}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_urban}
\caption{Percentile birth interval durations, sex ratios, and parity progression  
for urban women with 12 or more years of education by spell, sex composition, and period}
\label{fig:spacing_highest_urban}
\end{figure}

@@

\subsection{When Sex Selection Is Less Used}

To separate the effects of the introduction of sex selection and the other changes in
India, I first discuss how the lengths of birth intervals have changed in situations where 
sex selection is less used.
The group broadly covers women with no education, regardless of the sex composition of their 
children, and women with any education who have one or two sons already. 
Despite the lower level of sex selection, son preference is still evident with the 
shortest spacing when they have only girls. 
Notably, for those least likely to use sex selection---rural women with no education---the 
difference in birth interval length across sex compositions has grown over time as spacing when 
sons are present has become longer.

% [25th percentile birth intervals]
A remarkably high proportion of birth intervals are still very short.
For all but the most educated, 25\% or more have their second and third child 
within 24 months of the previous birth.
These intervals are substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Even with the more substantial lengthening of birth intervals for higher parities, the 25th
percentile duration is still around 24 months for the fourth spell for women with less than
eight years of education.

% [median birth intervals]
Median birth interval lengths have also increased relatively little---only three to six months 
over the four decades---compared to around 3.5 months \emph{per decade} in other countries with 
declining fertility \citep{Rutstein2011,Casterline2016}.%
\footnote{
The NFHS reports show median durations of closed birth intervals of approximately 31 months, 
which have barely moved over time, underscoring the importance of accounting for censoring 
when examining birth spacing.
}
The result is that most of the median birth interval durations are still at 36 months or
below, with the shortest only 29 months.

Birth intervals appeared to lengthen the most for women who are the least likely to work.
For example, from lowest to highest education, the average duration of third-spell birth 
intervals for urban women with one boy and one girl increased by 2.7, 3.4, 5.8, and 1.8 
months over the four decades.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0.
See the online appendix tables for the average birth interval durations.
}
Hence, women with the lowest labor force participation---those with 8--11 years 
of education---also saw the largest increases in average spacing, possibly driven by the 
substantial improvement in household income for this group from economic growth.

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth interval durations, 
where the more the parity progression probabilities declined, the more the birth interval 
lengthened. 
For example, the probability of a fourth birth for urban women with 8--11 years 
of education and two sons and one girl has declined by almost 40 percentage points as the 
75th percentile birth interval length increased by 22 months.
Compare this with rural women of no education with a boy as their first child, for whom the 
probability of a third birth declined by fewer than six percentage points while the birth 
interval increased only slightly over two months.

@@

These results are in line with prior research showing that falling fertility is associated 
with a lengthening of particularly the longer birth intervals, although why is still an 
unresolved question \citep{Casterline2016}.
The most educated women who already have a son is an exception to this trend; the 
probability of a third birth declined rapidly, but the birth interval lengths changed little. 
These women both have better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}. 
Hence, it is possible that the lengthening of birth intervals with falling fertility arises
because unintended births become a relatively larger proportion of the total number of
births for a given parity.

@@

A clear illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
With two girls, almost 80\% of the third births are boys, and the 75th percentile 
birth interval length is close to 70 months. 
This interval is about 13 months longer than if they had at least one son and represents 
an increase of almost 21 months over the four decades.

Even more striking is that most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These birth spacing changes may even be underestimated because this particular group
appears to have had access to sex selection even before it became widespread, as shown by
the unequal sex ratio for the 1972--1984 period for women with two girls.

The 75th percentile changes are the most dramatic, but sex selection also affects the 
25th and median birth interval lengths. 
For the best-educated urban women with two girls, the 25th percentile birth interval
length increased by six months, or 23\%, while the median percentile birth interval length
increased by 15 months (43\%).

@@

@@

Sex selection, however, is behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth interval lengths. 
Over the four decades, the median birth interval length for this group increased by almost 
15 months, 
and the 75th percentile birth interval length increased by a staggering 21 months, most of 
that within a decade.

@@

Consistent with a more substantial bias in the fertility rate when the age of marriage 
and the length of birth intervals increase, the absolute bias is least in the first 
and the last period and highest in the middle two periods.
Hence, the fertility rate declined too fast from the mid-1980s to the century's end.
Only recently, as the rate of increase for the birth interval lengths has slowed, have the 
two fertility measures begun to converge again.
Even with the convergence, the predicted 2005--2016 cohort fertility is still above the 
1992--1993 fertility rate for every group, except urban women with no education. 
Furthermore, for the last period, the predicted cohort fertility remains at least 
10\%--20\% higher than the fertility rate.

@@

Finally, even with the declines in the predicted cohort fertility, it is still mostly 
above replacement.
Only for urban women with 12 or more years of education is the predicted cohort
fertility clearly below 2.1 children.
Even then, cohort fertility is still more than 0.3 children higher than 
the fertility rate estimate of 1.5.
Furthermore, the predicted cohort fertility numbers are likely too low because I use only
the first four births and births before the imposed 105-month censoring of birth interval
durations.

@@

The dependent variable is whether the child died within the first 12 months of life.
The main set of explanatory variables consists of dummies for the spacing from
the prior birth.
The birth interval duration dummies cover 12-month periods, starting nine months after the 
prior birth, until the 57-month dummy, which covers until 105 months after the prior birth.
I use dummies for sex of the index child and the sex composition 
of the prior children.
The birth spacing dummies, the sex of index child, and the sex composition dummies
are all interacted.
Because the actual number of abortions is unobserved, the interactions between
the sex composition of prior children and the sex of the index child serve
as proxies for the use of sex selection.
The other explanatory variables are the same as above, and estimations
are done separately by education level and parity.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_2_low_med}
\caption{Infant mortality by preceding birth interval length across periods for second child of women with 
no education and women with 1--7 years of education}
\label{fig:mortality_low_med}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_2_high_highest}
\caption{Infant mortality by preceding birth interval length across periods for second child of women with 
8--11 and 12 and above years of education}
\label{fig:mortality_high_highest}
\end{figure}

@@

% What has happened to spacing and why
The most substantial lengthening of birth intervals came from the best-educated women
because of their substantial use of sex selection combined with falling fertility.
Take, for example, women with 12 or more years of education who had two girls. 
As the sex ratio increased to close to 80\% boys, the expected median birth interval length
increased by almost 15 months, and the 75th percentile interval length increased by 21 months.
Most of the increase in the long intervals came immediately after the introduction of sex 
selection in India.

@@

Tempo effects are studied extensively in the literature \citep[see, for example, ][]{Bongaarts1999}.
Still, there are, to my knowledge, no other cases where there has been as 
substantial an increase in birth interval lengths and associated bias in fertility rates 
as for India.
It is conceivable that we might see increases in the total fertility rate as birth spacing 
stabilizes or even shortens again if interventions against sex selection are successful.

@@

NFHS enumerators probe for any missed births, although the method depends on the survey.
NFHS-1 probe for each calendar birth interval that is four or more years.
NFHS-2 asked for stillbirths, spontaneous and induced abortions and also probed 
for each calendar birth interval four or more years.
NFHS-3 and NFHS-4 did not directly use birth interval lengths, but asked whether there were any 
other live births between (name of previous birth) and (name), including any children who 
died after birth, and asked for births before the birth listed as first birth and
after the last birth listed as the last birth.

@@

Figures \ref{fig:spacing_low_urban} and \ref{fig:spacing_highest_rural} show 25th, 50th,
and 75th percentile birth interval lengths, the sex ratio, and the probability of parity
progression by spell for urban women with no education and rural women with 12 or
more years of education, respectively.

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_low_urban}
\caption{Percentile birth interval lengths, sex ratios, and parity progression  
for urban women with no education by spell, sex composition, and period}
\label{fig:spacing_low_urban}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{bs_highest_rural}
\caption{Percentile birth interval lengths, sex ratios, and parity progression  
for rural women with 12 or more years of education by spell, sex composition, and period}
\label{fig:spacing_highest_rural}
\end{figure}

The first set of tables, Tables \ref{tab:p25_p50_p75_low}, \ref{tab:p25_p50_p75_med},
\ref{tab:p25_p50_p75_high}, and \ref{tab:p25_p50_p75_highest} show 25th, 50th, and
75th percentile birth interval lengths together with their standard errors.
The standard errors for all measures are based on bootstrapping,
where the model is repeatedly estimated using resampling with replacement.

The second set of tables, 
Tables \ref{tab:avg_sex_ratio_low}, \ref{tab:avg_sex_ratio_med}, 
\ref{tab:avg_sex_ratio_high}, and \ref{tab:avg_sex_ratio_highest}, show predicted average 
birth interval lengths, sex ratios, and probabilities of having a birth by decade, spell, and sex 
composition for the four education levels separated by the area of residence, together 
with bootstrapped standard errors for all three outcomes.
To find the average birth interval length, I calculate, for each woman, the probability of 
giving birth in each $t$, and her expected spell length from these probabilities. 
I then average the individual expected spell lengths across women using their parity 
progression probabilities as weights. Finally, I add nine months because spells begin 
nine months after the previous birth.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_3_low_med}
\caption{Infant mortality by preceding birth interval length across periods for third child of women with 
no education and women with 1--7 years of education}
\label{fig:mortality_low_med_spell_3}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio=true]{mortality_spell_3_high_highest}
\caption{Infant mortality by preceding birth interval length across periods for third child of women with 
8--11 and 12 and above years of education}
\label{fig:mortality_high_highest_spell_3}
\end{figure}
